Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2025-04-28 Paper/paper.tex
Added about beta reducability

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{mdframed}
%\usepackage{cleveref}

@@

\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
\begin{document}

@@

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (we don't require $\Sigma,\Gamma$ to be disjoint),
    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
    \item otherwise $\beta(\xi)=\beta'(\xi)$.
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

@@

According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.

@@

\ari{The following section is just me writing down some ideas, far from finished}

\subsection{Multi-$\beta$-Reducible Languages}

For $n\geq1$, a $\BR(n)$-schema is a tuple $(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\set{\hb_i}_{i=0}^{n-1})$ where $\Sigma,\Gamma,\square,\blacksquare,\bullet,s$ are as before and now
\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n-1} \]
For each $i\in\set{0,\dots,n-1}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.

We define \MBR{} to be $\MBR=\bigcup_{n\geq1}\BR(n)$.

This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
We construct a $\BR(5)$-schema as follows:
\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
Where we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& \square,A\to A',1\\
        & \square,B\to B',3\\
    \hb_1\colon\quad& A,\blacksquare\to A',2\\
    \hb_2\colon\quad& A',A'\to\bullet,2\\
        &A',\epsilon\to\epsilon,2\\
        &A',\blacksquare\to A',0\\
    \hb_3\colon\quad& B,\blacksquare\to B',2\\
    \hb_4\colon\quad& B',B'\to\bullet,2\\
        &B',\epsilon\to\epsilon,2\\
        &B',\blacksquare\to B',0\\
\end{align*}
Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
Otherwise delete them both.

For the language of palindromes $\set{w\in\set{a,b}^*}[w=w^r]$ we can define a $\BR(3)$-schema where $\bullet=\epsilon$ and
\begin{align*}
    \hb_0\colon\quad& \square,A\to\epsilon,1\\
        & \square,B\to\epsilon,2\\
    \hb_1\colon\quad& A,\blacksquare\to\epsilon,0\\
    \hb_2\colon\quad& B,\blacksquare\to\epsilon,0\\
\end{align*}

But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
Indeed, we can construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
And we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& B,A\to X,0\\
        &C,B\to X,0\\
        &C,\blacksquare\to C,1\\
    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
    \hb_3\colon\quad& C,\blacksquare\to\bullet,3\\
        &C,\epsilon\to\epsilon,1
\end{align*}
Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.

>>>>

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
%\usepackage{mdframed}
%\usepackage{cleveref}

@@

%\newcommand{\yoni}[1]{{\begin{mdframed}[linecolor=red]{\color{red}#1}\end{mdframed}}}
%\newcommand{\ari}[1]{{\begin{mdframed}[linecolor=blue]{\color{blue}#1}\end{mdframed}}}
\def\yoni#1{\par\noindent\fbox{\color{red}#1}\par}
\def\ari#1{\par\noindent\fbox{\color{blue}#1}\par}
\begin{document}

@@

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (we don't require $\Sigma,\Gamma$ to be disjoint),
    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon(\Gamma\cup\set\square)\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
    \item otherwise $\beta(\xi)=\beta'(\xi)$.
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

@@

According to theorem \gobble\ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.

@@

\subsection{Variants of \BR}

In this section we define variants of $\beta$-reducible languages, which are all special cases (including the initial variant
already defined) of the following definition:

    A \emph{nondeterministic $n$-\BR{} schema} (or an $n$-\NBR{} schema), for $n\in\bN$, is a tuple
    $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ (like before), where
    \begin{enumerate}
        \item $\Sigma$ is a finite alphabet of terminals;
        \item $\Gamma$ is a finite alphabet of non-terminals (not necessarily disjoint from $\Sigma$);
        \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$;
        \item $\bullet$ is a symbol (perhaps in $\Sigma$ or $\Gamma$), or the empty word;
        \item $s$ is a state, which is a map $\Sigma\to\Gamma$;
        \item $\hb$ is an initial $\beta$-reducer, which is a function
        $$ \hb\colon(\Gamma\cup\set\square)\times(\Gamma\cup\set{\epsilon,\blacksquare})\to
        \P((\Gamma\cup\set{\epsilon,\bullet})^n\cup\set\perp) $$
    \end{enumerate}

    Given a nondeterministic $n$-\NBR{} schema $B$, we define the function
    $\beta'\colon(\Gamma\cup\set\bullet)^*\to\P((\Gamma\cup\set\bullet)^*)$ as follows:
    \begin{enumerate}
        \item $\beta'(\bullet)=\set\bullet$;
        \item $\beta'(\sigma)=\hb(\sigma,\blacksquare)$ if defined, otherwise $\set\sigma$;
        \item if $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item if $\perp\notin\hb(\sigma_1,\sigma_2)$, $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$
            (where the concatenation of sets is defined word-wise) if defined, and if $\perp$ is in the reduction, then take
            the union of this with $\sigma_1\
            otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\xi'$ if defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
    \end{enumerate}
    Then we define the function $\beta\colon(\Gamma\cup\set\bullet)^*\to\P((\Gamma\cup\set\bullet)^*)$ as follows:
    \begin{enumerate}
        \item $\beta(\epsilon)=\hb(\square,\blacksquare)$ if defined, otherwise $\set\epsilon$;
        \item $\beta(\sigma\xi)=\hb(\square,\sigma)\xi$ if defined, otherwise:
        \item $\beta(\xi)=\beta'(\xi)$.
    \end{enumerate}
    This naturally extends to a function $\beta\colon\P((\Gamma\cup\set\bullet)^*)\to\P((\Gamma\cup\set\bullet)^*)$.
    We then define the \emph{language of $B$} to be
    $$ \L(B) = \set{w\in\Sigma^*}[\bullet\in\beta^k(s(w))\hbox{ for some $k\geq0$}] $$

The class $n$-\NBR{} is defined to be the class of all languages recognized by a nondeterministic $n$-\BR{} schema.
Then we define \sNBR{} to be the union of all $n$-\NBR{} for $n\geq1$.

    An $n$-\NBR{} schema is \emph{deterministic} if
